Reanalysis Step 5: Outcome Analysis with Three-Way Interaction

### 2025-09-04 reanalysis-5-outcome_analysis-interaction.qmd
shared test

```{r joint-tests}
pool_chi_square_tests <- function(models, hypothesis_strings, test_name) {

  # Perform test on each imputation
  test_stats <- numeric(length(models))
  df_tests <- numeric(length(models))

  for (i in 1:length(models)) {
    test_result <- linearHypothesis(
      models[[i]],
      hypothesis_strings,
      vcov = vcov(models[[i]])
    )
    test_stats[i] <- test_result$Chisq[2]
    df_tests[i] <- test_result$Df[2]
  }

  # Pool using Li, Raghunathan, and Rubin (1991) method for chi-square tests
  # D-bar: average test statistic
  D_bar <- mean(test_stats)
  df <- df_tests[1]  # Should be same across all imputations

  # For large samples, can use D_bar directly with original df
  # For more accuracy, would need to compute the adjusted statistic
  p_value <- pchisq(D_bar, df = df, lower.tail = FALSE)

  return(list(
    test_name = test_name,
    D_bar = D_bar,
    df = df,
    p_value = p_value
  ))
}

first_model <- models[[1]]

# Test 1: Joint test of three-way interaction
threeway_terms_test <- grep("ever_lapse.*:dr_severity.*:any_treatment", 
                            names(coef(first_model)), value = TRUE)

# Test 1: Three-way interaction
if (length(threeway_terms_test) > 0) {
  hypothesis_3way <- paste0(threeway_terms_test, " = 0")
  pooled_test_3way <- pool_chi_square_tests(models, hypothesis_3way, "Three-way interaction")
}

# Test 2: Joint test of lapse × DR interaction (averaging over treatment)
lapse_dr_terms <- grep("ever_lapse.*:dr_severity(?!.*:any_treatment)", 
                      names(coef(first_model)), value = TRUE, perl = TRUE)

if (length(lapse_dr_terms) > 0) {
  hypothesis_lapse_dr <- paste0(lapse_dr_terms, " = 0")
  pooled_test_lapse_dr <- pool_chi_square_tests(models, hypothesis_lapse_dr, "Lapse × DR interaction")
}

# Test 3: Joint test of lapse × treatment interaction (averaging over DR)
lapse_tx_terms <- grep("ever_lapse.*:any_treatment(?!.*:dr_severity)", 
                      names(coef(first_model)), value = TRUE, perl = TRUE)

if (length(lapse_tx_terms) > 0) {
  hypothesis_lapse_tx <- paste0(lapse_tx_terms, " = 0")
  pooled_test_lapse_tx <- pool_chi_square_tests(models, hypothesis_lapse_tx, "Lapse × treatment interaction")
}
```

@@

```{r joint-test-results}
# Compile all test results
joint_tests <- data.frame(
  Test = character(),
  Chi_Square = numeric(),
  df = numeric(),
  P_Value = character(),
  Significant = character(),
  stringsAsFactors = FALSE
)

if (exists("pooled_test_3way")) {
  joint_tests <- rbind(joint_tests, data.frame(
    Test = "Lapse × DR × Treatment (3-way)",
    Chi_Square = round(pooled_test_3way$D_bar, 3),
    df = pooled_test_3way$df,
    P_Value = format.pval(pooled_test_3way$p_value, digits = 4, eps = 0.0001),
    Significant = ifelse(pooled_test_3way$p_value < 0.05, "Yes", "No")
  ))
}

if (exists("pooled_test_lapse_dr")) {
  joint_tests <- rbind(joint_tests, data.frame(
    Test = "Lapse × DR Severity",
    Chi_Square = round(pooled_test_lapse_dr$D_bar, 3),
    df = pooled_test_lapse_dr$df,
    P_Value = format.pval(pooled_test_lapse_dr$p_value, digits = 4, eps = 0.0001),
    Significant = ifelse(pooled_test_lapse_dr$p_value < 0.05, "Yes", "No")
  ))
}

if (exists("pooled_test_lapse_tx")) {
  joint_tests <- rbind(joint_tests, data.frame(
    Test = "Lapse × Treatment",
    Chi_Square = round(pooled_test_lapse_tx$D_bar, 3),
    df = pooled_test_lapse_tx$df,
    P_Value = format.pval(pooled_test_lapse_tx$p_value, digits = 4, eps = 0.0001),
    Significant = ifelse(pooled_test_lapse_tx$p_value < 0.05, "Yes", "No")
  ))
}

kable(joint_tests, caption = "Pooled joint hypothesis tests (Li-Raghunathan-Rubin method)")
```

>>>>

```{r joint-tests}
# Function to pool Wald tests across multiple imputations using Rubin's rules
pool_wald_tests <- function(models, term_indices, test_name) {
  n_imp <- length(models)
  n_params <- length(term_indices)

  # Storage for estimates and covariances from each imputation
  beta_list <- list()
  vcov_list <- list()

  # Extract coefficients and vcov for specified terms from each imputation
  for (i in 1:n_imp) {
    coef_full <- coef(models[[i]])
    vcov_full <- vcov(models[[i]])

    beta_list[[i]] <- coef_full[term_indices]
    vcov_list[[i]] <- vcov_full[term_indices, term_indices, drop = FALSE]
  }

  # Rubin's rules pooling
  # Step 1: Pool point estimates (beta_bar)
  beta_bar <- Reduce("+", beta_list) / n_imp

  # Step 2: Within-imputation variance (U_bar)
  U_bar <- Reduce("+", vcov_list) / n_imp

  # Step 3: Between-imputation variance (B)
  beta_matrix <- do.call(rbind, beta_list)
  B <- cov(beta_matrix)

  # Step 4: Total variance (T)
  T_var <- U_bar + (1 + 1/n_imp) * B

  # Step 5: Wald statistic
  # Check if T_var is invertible
  T_inv <- try(solve(T_var), silent = TRUE)
  if (inherits(T_inv, "try-error")) {
    warning(paste0("Covariance matrix not invertible for ", test_name))
    return(list(
      test_name = test_name,
      F_stat = NA,
      df1 = n_params,
      df2 = NA,
      p_value = NA
    ))
  }

  wald_stat <- as.numeric(t(beta_bar) %*% T_inv %*% beta_bar)

  # Step 6: Reference distribution with Barnard-Rubin adjustment
  # Relative increase in variance
  r <- mean(diag((1 + 1/n_imp) * B %*% solve(U_bar)))

  # Barnard-Rubin degrees of freedom
  df_m <- (n_imp - 1) * (1 + 1/r)^2

  # For multi-parameter test, use F distribution
  F_stat <- wald_stat / n_params
  df1 <- n_params
  df2 <- df_m

  # P-value from F distribution
  p_value <- pf(F_stat, df1, df2, lower.tail = FALSE)

  return(list(
    test_name = test_name,
    F_stat = F_stat,
    df1 = df1,
    df2 = round(df2),
    p_value = p_value
  ))
}

first_model <- models[[1]]
all_coef_names <- names(coef(first_model))

# Test 1: Three-way interaction terms only
threeway_indices <- which(grepl("ever_lapse.*:dr_severity.*:any_treatment", all_coef_names))

if (length(threeway_indices) > 0) {
  log_message(paste0("Testing three-way interaction with ", length(threeway_indices), " parameters"))
  pooled_test_3way <- pool_wald_tests(models, threeway_indices, "Three-way interaction")
}

# Test 2: All interaction terms (2-way and 3-way jointly)
all_interaction_patterns <- c(
  "ever_lapse.*:dr_severity",
  "ever_lapse.*:any_treatment", 
  "dr_severity.*:any_treatment"
)

all_interaction_indices <- which(grepl(paste(all_interaction_patterns, collapse = "|"), all_coef_names))

if (length(all_interaction_indices) > 0) {
  log_message(paste0("Testing all interactions jointly with ", length(all_interaction_indices), " parameters"))
  pooled_test_all <- pool_wald_tests(models, all_interaction_indices, "All interactions (2-way + 3-way)")
}

# Test 3: Lapse × DR interaction only (2-way, excluding 3-way)
lapse_dr_indices <- which(grepl("ever_lapse.*:dr_severity", all_coef_names) & 
                         !grepl("any_treatment", all_coef_names))

if (length(lapse_dr_indices) > 0) {
  log_message(paste0("Testing lapse × DR interaction with ", length(lapse_dr_indices), " parameters"))
  pooled_test_lapse_dr <- pool_wald_tests(models, lapse_dr_indices, "Lapse × DR interaction")
}

# Test 4: Lapse × treatment interaction only (2-way, excluding 3-way)
lapse_tx_indices <- which(grepl("ever_lapse.*:any_treatment", all_coef_names) & 
                         !grepl("dr_severity", all_coef_names))

if (length(lapse_tx_indices) > 0) {
  log_message(paste0("Testing lapse × treatment interaction with ", length(lapse_tx_indices), " parameters"))
  pooled_test_lapse_tx <- pool_wald_tests(models, lapse_tx_indices, "Lapse × treatment interaction")
}

# Test 5: DR × treatment interaction only (2-way, excluding 3-way)
dr_tx_indices <- which(grepl("dr_severity.*:any_treatment", all_coef_names) & 
                      !grepl("ever_lapse", all_coef_names))

if (length(dr_tx_indices) > 0) {
  log_message(paste0("Testing DR × treatment interaction with ", length(dr_tx_indices), " parameters"))
  pooled_test_dr_tx <- pool_wald_tests(models, dr_tx_indices, "DR × treatment interaction")
}
```

@@

```{r joint-test-results}
# Compile all test results
joint_tests <- data.frame(
  Test = character(),
  F_Statistic = numeric(),
  Num_DF = numeric(),
  Den_DF = numeric(),
  P_Value = character(),
  Significant = character(),
  stringsAsFactors = FALSE
)

# Add three-way interaction test
if (exists("pooled_test_3way") && !is.na(pooled_test_3way$F_stat)) {
  joint_tests <- rbind(joint_tests, data.frame(
    Test = "Lapse × DR × Treatment (3-way only)",
    F_Statistic = round(pooled_test_3way$F_stat, 4),
    Num_DF = pooled_test_3way$df1,
    Den_DF = pooled_test_3way$df2,
    P_Value = format.pval(pooled_test_3way$p_value, digits = 4, eps = 0.0001),
    Significant = ifelse(pooled_test_3way$p_value < 0.05, "Yes", "No")
  ))
}

# Add all interactions test
if (exists("pooled_test_all") && !is.na(pooled_test_all$F_stat)) {
  joint_tests <- rbind(joint_tests, data.frame(
    Test = "All interactions (2-way + 3-way)",
    F_Statistic = round(pooled_test_all$F_stat, 4),
    Num_DF = pooled_test_all$df1,
    Den_DF = pooled_test_all$df2,
    P_Value = format.pval(pooled_test_all$p_value, digits = 4, eps = 0.0001),
    Significant = ifelse(pooled_test_all$p_value < 0.05, "Yes", "No")
  ))
}

# Add two-way interaction tests
if (exists("pooled_test_lapse_dr") && !is.na(pooled_test_lapse_dr$F_stat)) {
  joint_tests <- rbind(joint_tests, data.frame(
    Test = "Lapse × DR (2-way only)",
    F_Statistic = round(pooled_test_lapse_dr$F_stat, 4),
    Num_DF = pooled_test_lapse_dr$df1,
    Den_DF = pooled_test_lapse_dr$df2,
    P_Value = format.pval(pooled_test_lapse_dr$p_value, digits = 4, eps = 0.0001),
    Significant = ifelse(pooled_test_lapse_dr$p_value < 0.05, "Yes", "No")
  ))
}

if (exists("pooled_test_lapse_tx") && !is.na(pooled_test_lapse_tx$F_stat)) {
  joint_tests <- rbind(joint_tests, data.frame(
    Test = "Lapse × Treatment (2-way only)",
    F_Statistic = round(pooled_test_lapse_tx$F_stat, 4),
    Num_DF = pooled_test_lapse_tx$df1,
    Den_DF = pooled_test_lapse_tx$df2,
    P_Value = format.pval(pooled_test_lapse_tx$p_value, digits = 4, eps = 0.0001),
    Significant = ifelse(pooled_test_lapse_tx$p_value < 0.05, "Yes", "No")
  ))
}

if (exists("pooled_test_dr_tx") && !is.na(pooled_test_dr_tx$F_stat)) {
  joint_tests <- rbind(joint_tests, data.frame(
    Test = "DR × Treatment (2-way only)",
    F_Statistic = round(pooled_test_dr_tx$F_stat, 4),
    Num_DF = pooled_test_dr_tx$df1,
    Den_DF = pooled_test_dr_tx$df2,
    P_Value = format.pval(pooled_test_dr_tx$p_value, digits = 4, eps = 0.0001),
    Significant = ifelse(pooled_test_dr_tx$p_value < 0.05, "Yes", "No")
  ))
}

if (nrow(joint_tests) > 0) {
  kable(joint_tests, 
        caption = "Pooled Wald Tests for Interaction Terms (Rubin's Rules with Barnard-Rubin DF)",
        col.names = c("Test", "F", "df1", "df2", "P-value", "Sig."))
} else {
  cat("No interaction tests could be performed.\n")
}
```
